Sampling requirements to detect ecosystem change

### 2020-02-05 R/manuscript_draft.Rmd
modified theory figs and added text

Authors: \textsc{Rosalie Bruel$^{1}$ and Easton R. White$^{2}$}
\vspace{5 mm}

Addresses: \normalsize{\indent $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA \\ $^{2}$Department of Biology, University of Vermont, VT 05405, USA}
\end{center}

@@

We explored how each of these model parameters affected our ability to detect a change point. We simulate an entire time series (the "true" data for comparison), but we also examine the effect of not sampling every year. We specifically study how the amount of samples included and the type of sampling affects the detection probability.

```{r theoretical_results, fig.cap='Theoretical results here detailing parameter sensitivity - we should make this about sample size instead of probability correct',message=F,warning=F,echo=F,eval=F}

source('R/sampling_regular.R')
source('R/build_time_series.R')
source('R/updated_param_sensitivity.R')

@@

@@

In line with theory about monitoring, we found that the probability of correcting identifying a changepoint decreased with larger levels of population variability ($\sigma$) and autocorrelation ($\phi$). We also found that the probability of correct changepoint detection increases with larger shift sizes.

The probability of correcting identifying a changepoint, of course, depends strongly on the number of samples.

@@

# Discussion

@@

Limits: Systematic sampling of a time-series is required to detect critical slowing-down or early warning of shifts (ref, maybe among others Frossard et al 2015, Doncaster et al 2016). The lower number of input data prior to ordination did not impact the finding of the main transition. However, if other metrics are of interest to the investigators, such as species richness, a more systematic or at least more extensive approach would be require.

>>>>

\textsc{Rosalie Bruel$^{1}$ and Easton R. White$^{2}$}
\vspace{5 mm}

\normalsize{\indent $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA \\ $^{2}$Department of Biology, University of Vermont, VT 05405, USA}
\end{center}

@@

We explored how each of these model parameters affected our ability to detect a change point. We simulate an entire time series (the "true" data for comparison), but we also examine the effect of not sampling every year. We specifically study how the amount of samples included and the type of sampling affects the detection probability. For simulations, statistical power is the fraction of simulations that were able to detect a changepoint. We define a detection as an estimate within five time points (over 100 time points) of the true changepoint. We define $T_{min}$ as the minimum number of samples required to achieve 0.8 statistical power.

```{r theoretical_results,message=F,warning=F,echo=F,eval=F}

source('R/sampling_regular.R')
source('R/sampling_random.R')
source('R/build_time_series.R')
source('R/updated_param_sensitivity.R')

@@

```{r theoretical_results_min_time,message=F,warning=F,echo=F,eval=F}

source('R/sampling_regular.R')
source('R/build_time_series.R')
source('R/updated_param_sensitivity.R')
source('R/calc_min_time2.R')
require(tidyverse)

# Sigma results
sigma_vec = seq(0.1,30,4)
min_time  = rep(0,times=length(sigma_vec))
sigma_output <- data.frame(sigma_vec,min_time)

for (param_num in 1:nrow(sigma_output)){
  sigma_output$min_time[param_num] <- calc_min_time('sigma',sigma_output$sigma_vec[param_num])
  #print(sigma_output)
}

write.csv(x = sigma_output,file = 'Output/theoretical_results_sigma.csv',quote=FALSE)

# Phi results

phi_vec = seq(-0.6,0.6,0.2)
min_time  = rep(0,times=length(phi_vec))
phi_output <- data.frame(phi_vec,min_time)

for (param_num in 1:nrow(phi_output)){
  phi_output$min_time[param_num] <- calc_min_time('phi',phi_output$phi_vec[param_num])
  #print(phi_output)
}

write.csv(x = phi_output,file = 'Output/theoretical_results_phi.csv',quote=FALSE)

# Shift size

shift_size_vec = rev(seq(0.1,0.9,0.05))
min_time  = rep(0,times=length(shift_size_vec))
shift_size_output <- data.frame(shift_size_vec,min_time)

for (param_num in 1:nrow(shift_size_output)){
  shift_size_output$min_time[param_num] <- calc_min_time('shift_size',shift_size_output$shift_size_vec[param_num])
  #print(shift_size_output)
}

write.csv(x = shift_size_output,file = 'Output/theoretical_results_shift_size.csv',quote=FALSE)

```{r theoretical_results_min_time_plot, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size. The default parameters are $\\sigma = 3$, $\\phi = 0.5$, and a shift size $=0.75$. The exact timing of the true changepoint varied for each simulation between time steps 30 and 70. \\label{fig:theoretical_min_time}',message=F,warning=F,echo=F,eval=T}

par(mfrow=c(1,3))

sigma <- read.csv('Output/theoretical_results_sigma.csv',header=TRUE)
plot(sigma$sigma_vec,sigma$min_time,ylab='Minimum number of samples required',xlab='sigma',pch=16,cex.lab=1.3,las=1,ylim=c(0,40),col=1)

# Phi results

phi <- read.csv('Output/theoretical_results_phi.csv',header=TRUE)
plot(phi$phi_vec,phi$min_time,ylab=' ',xlab='phi',pch=16,cex.lab=1.3,las=1,ylim=c(0,40),col=1)

# Shift size results

shift_size <- read.csv('Output/theoretical_results_shift_size.csv',header=TRUE)
plot(shift_size$shift_size_vec,shift_size$min_time,ylab=' ',xlab='shift size',pch=16,cex.lab=1.3,las=1,ylim=c(0,40),col=1)

@@

In line with theory about monitoring, we found that the probability of correcting identifying a changepoint decreased with smaller levels of population variability ($\sigma$) and autocorrelation ($\phi$). We also found that the probability of correct changepoint detection increases with larger shift sizes, which is essentially the effect size. Thus, the number of sample required to obtain high statistical power (above 0.8) increased with larger population variability, lower autocorrelation, and smaller shift sizes (Fig. \ref{fig:theoretical_min_time}).

@@

# General Discussion

Outline:

- theoretical results
- example with the case study
- how this fits in with other work
- limitations and future work

@@

Limitations:

- It is not limited to paleo data
- Depends on the question being asked
- Specific number of samples depends on the system dynamics

Systematic sampling of a time-series is required to detect critical slowing-down or early warning of shifts (ref, maybe among others Frossard et al 2015, Doncaster et al 2016). The lower number of input data prior to ordination did not impact the finding of the main transition. However, if other metrics are of interest to the investigators, such as species richness, a more systematic or at least more extensive approach would be require.

Concluding paragraph:
